Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-12 concl.tex
note to concl
synching everything in case I forgot something

\section{ Conclusions \& Future Work}\label{sec:conc}

\begin{enumerate}
\item  This need to be rewritten to conclude and address this paper...
\end{enumerate}

The results presented here suggest that permutation entropy---a ordinal
calculation of forward information transfer in a time series---is an effective
metric for predictability of computer performance traces. Experimentally, traces
with a persistent PE $\gtrapprox 0.97$ have a natural level of complexity that
may overshadow the inherent determinism in the system dynamics, whereas traces
with PE $\lessapprox 0.7$ seem to be highly predictable (viz., at least an order
of magnitude improvement in nRMSPE).Further, the persistent WPE values of 0.5--
0.6 for the {\tt col\_major} trace are consistent with dynamical chaos, further
corroborating the results of~\cite{mytkowicz09}.

If information is the limit, then gathering and using more information is an
obvious next step.  There is an equally obvious tension here between data length
and prediction speed: a forecast that requires half a second to compute is not
useful for the purposes of real-time control of a computer system with a MHz
clock rate.  Another alternative is to sample several system variables
simultaneously and build multivariate delay-coordinate embeddings.  Existing
approaches to that are computationally prohibitive
\cite{cao-multivariate-embedding}.  We are working on alternative
methods that sidestep that complexity.

%%This use of permutaiotn entropy is a useful application of information
%%theory to computer performance modeling because it gives you a simple
%%and fast way to decide whether building a model and trying to forecast
%%the future is worthwhile or if guessing the mean is just as effective.
%%End with a sentence tying back to power management and world peace.

%\begin{it}
%Summarize the results.

@@

>>>>

\section{ Conclusions \& Future Work }\label{sec:conc}

It's obvious that no forecast method can work if a time series has no
predictive structure, but it's hard to define or measure predictive
structure.  In this paper, we have argued that redundancy---insert
short definition here---is a useful definition of predictive structure
in a time series, and that weighted permutation entropy is an
effective way to estimate how much of that structure is present in a
given real-valued, scalar time series.  Even if predictive structure
exists, a given forecast method may not work---simply because it
cannot capture and/or exploit the structure that is present (e.g., a
linear model of a nonlinear process).  This fairly obvious point can
be a serious practical challenge.  When a model isn't working, the
practitioner doesn't know whether it's because there's no predictive
structure (i.e., no model will work) or whether the model simply isn't
good enough.  Fortunately, weighted permutation entropy allows us to
make this distinction for noisy real-valued time series.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
successfully predict.  We have argued that
redundancy~\cite{crutchfield2003} is a useful definition of the
inherent predictability of a time series.  To operationalize that
definition, we use an approximation of the Kolmogorov-Sinai
entropy~\cite{lind95}, estimated using a weighted version of the
permutation entropy of~\cite{bandt2002per}.  This WPE technique---an
ordinal calculation of forward information transfer in a time
series---is ideal for our purposes because it works with real-valued
data and is known to converge to the true entropy value.  Using a
variety of forecast models and more than a hundred time-series data
sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: summarize \& discuss here.

% Further, the persistent WPE values of 0.5--0.6 for the {\tt
%   col\_major} trace are consistent with dynamical chaos, further
% corroborating the results of~\cite{mytkowicz09}.

An important practical corollary to this empirical correlation of
predictability and WPE is a strategy for assessing appropriateness of
forecast methods.  If the forecast produced by a particular method is
poor but the time series contains a significant amount of predictive
structure, one can reasonably conclude that that method is inadequate
to the task and that one should seek another method.  The nonlinear
LMA method, for instance, performs better in most cases because it is
more general.  The \naive ~method, which simply predicts the mean,
works better in noisy signals because it's a filter.  The dashed line
in Figures blah is a useful empirical guideline for knowing when a
model is not well-matched to the task at hand: predictions whose MASE
scores are more than foo times their WPE are an indication that the
time series has more predictive structure than that method can capture
and exploit.

If information is the limit, then gathering and using more information
is an obvious next step.  There is an equally obvious tension here
between data length and prediction speed: a forecast that requires
half a second to compute is not useful for the purposes of real-time
control of a computer system with a MHz clock rate.  Another
alternative is to sample several system variables simultaneously and
build multivariate delay-coordinate embeddings.  Existing approaches
to that (e.g., \cite{cao-multivariate-embedding,deyle-sugihara2011})
are computationally prohibitive, however.  We are working on
alternative methods that sidestep that complexity.

@@

Nonstationarity is a serious challenge in any time-series modeling
problem.  One of the most common uses of permutation entropy is in
regime-shift detection: e.g., to detect seizures in brainwave data
\cite{cao2004det}.  The SVD signal in Figure~\ref{fig:what} was useful
here because it contains a number of different regimes.  We segmented
this signal visually, but one could imagine using permutation entropy
to do so instead.  This would be a win because it would be automatic,
which would move towards an adaptive modeling strategy, wherein one
would toss the old model at the regime boundary, measure WPE, then use
the result to choose which kind of model to build for the next regime.
More here from Ryan and Joshua.

Experiments here are from one system, but should work on others.  Give
some examples.  Close with mention of what that could enable (viz.,
adaptive models like ones at end of prev paragraph used as the core of
controllers).

% The way structure, information and complexity are processed by a
% generating process play a crucial role in the success of a given
% prediction scheme.
